Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-09-16 review/response.Rmd

---
title: "Response"
author: "H. Sherry Zhang"
date: "22/08/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\textcolor{red}{We really appreciate the helpful comments and feedback. Below is a point-by-point explanation of how we have changed the paper. The reviewer comments are in black and our responses are in red.}

\textcolor{red}{In our revision, we have used the new format for the journal and produced both a pdf format and an html format. The html format includes interactive graphics whenever appropriate.}

@@

### Major comment

-   "Sometimes, two optimisers can start with the same basis but finish with bases of opposite sign...However, this creates difficulties for comparing the optimisers since the end bases will be symmetric to the origin. A sign flipping step is conducted..." and the "Reconciling the orientation" section

    -   For 1D projections, flipping the sign is sufficient. For p-D projections of $\mathbb{R}^n$​, one may need a little more. To my knowledge most index functions are invariant to any rotations that happens within the p-subspace. This makes me wonder that in general, one may want to align two bases so that their difference does not contain any "within-subspace" components. In other words, one may want to compare representations in the Grassmannian manifold (space of k-subspaces) instead of in the Stiefel manifold (space of orthonormal k-frames).

\textcolor{red}{We agree that most index functions should be rotation invariant (this is in fact a desirable property, but there are exceptions where indexes can still be very useful but not rotation invariant, see discussion in Laa, Cook (2020)). Thus when comparing frames defining projections to d>=2 dimensions, we may want to introduce a rotation that will align the bases as much as possible within the common subspace, while potentially quantifying e.g. the angle between two planes. This is an interesting problem, however the implementation of this general approach is beyond the scope of the current paper. The text in section "Reconciling the orientation" has been updated to reflect this explanation.}

@@

### Minor comment

-   The software seems to be specific to projection pursuit. I wish the authors could comment more on its potential generality to other optimizations. For example, can one generalizes the tour method to visualize any parameter trace of an optimizer?

\textcolor{red}{It is feasible to generalise the tour method for diagnosing other optimisers, given that the parameter space can be simulated. In our examples, the orthonormality constraint regulates the parameter space of the projection basis ($p \times d$) to be an p-sphere, which can be simulated by the geozoo package. For other optimisers, reserachers need first to create a data collection object in a tibble that tracks the parameter during the optimisation, and then to use ferrn to produce the diagnostic plots as shown in the paper. Some examples of the data collection objects that we created are included in the ferrn package and should be referred to as the format for inputs into the explore_* functions. we have added a sentence in the conclusion to point to this possibility. }

### 2021-09-16 paper/zhang-cook-laa-langrene-menendez.Rmd
update web article
changes made during the meeting

---
title: "Visual Diagnostics for Constrained Optimisation with Application to Guided Tours"
author:
  - name: H.Sherry Zhang
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics 
    url: Melbourne, Australia
    email:  huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    url: Melbourne, Australia
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    address: Institute of Statistics
    url: Vienna, Austria
    email:  ursula.laa@boku.ac.at  
  - name: Nicolas Langrené
    affiliation: CSIRO Data61 
    address: 34 Village Street, Docklands VIC 3008 Australia
    url: Melbourne, Australia
    email: nicolas.langrene@csiro.au
  - name: Patricia Menéndez
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    url: Melbourne, Australia
    email:  patricia.menendez@monash.edu 
abstract: >
  A guided tour helps to visualise high-dimensional data by showing low-dimensional projections along a projection pursuit optimisation path. Projection pursuit is a generalisation of principal component analysis, in the sense that different indexes are used to define the interestingness of the projected data. While much work has been done in developing new indexes in the literature, less has been done on understanding the optimisation. Index functions can be noisy, might have multiple local maxima as well as an optimal maximum, and are constrained to generate orthonormal projection frames, which complicates the optimization. In addition, projection pursuit is primarily used for exploratory data analysis, and finding the local maxima is also useful. The guided tour is especially useful for exploration, because it conducts geodesic interpolation connecting steps in the optimisation and shows how the projected data changes as a maxima is approached. This work provides new visual diagnostics for examining a choice of optimisation procedure, based on the provision of a new data object which collects information throughout the optimisation. It has helped to diagnose and fix several problems with projection pursuit guided tour. This work might be useful more broadly for diagnosing optimisers, and comparing their performance. The diagnostics are implemented in the R package [ferrn](https://github.com/huizezhang-sherry/ferrn).
keywords: graphics, multivariate, optimization
preamble: >
  \usepackage{mathtools}
  \usepackage[ruled, linesnumbered]{algorithm2e}
output: 
  rjtools::rjournal_pdf_article
bibliography: zhang-cook-laa-langrene-menendez.bib
---

@@

```{r toy-pca-animated-interactive,  out.wdith= "100px", fig.cap = "Animated version of Figure \\ref{fig:toy-pca}. With animation, the progression of the search paths from start to finish can better identified and CRS (green) finishes the optimisation quicker than PD (brown).", eval=knitr::is_html_output()}
```

@@

One interesting situation observed in the previous examples is that, for some simulations, as shown on the left panel of Figure \ref{fig:flip-sign}, the target basis is generated on the other half of the basis space and the interpolator seems to draw a straight line to interpolate. Bases with opposite signs do not affect the projection and index value, but we would prefer the target to have the same orientation as the current basis. The orientation of two bases can be checked via calculating the determinant and a negative value suggests the two bases having different orientation. For 1D bases, this can be corrected by flipping the sign in one of the basis while the correction for a general p-D bases is more complex and beyond the scope of this paper. Here an orientation check is carried out once a new target basis is generated and the sign in the target basis will be flipped if a negative determinant is obtained. The interpolation after implementing the orientation check is shown on the right panel of Figure \ref{fig:flip-sign} where the unsatisfactory interpolation no longer exists.

@@

This paper has provided several visual diagnostics that can be used for diagnosing a complex optimisation procedure. The methods were illustrated using the optimisers available for projection pursuit guided tour. Here the constraint is the orthonormality condition of the projection bases. The approach used broadly applies to other constrained optimisers.

The tools were implemented in the ferrn package and can track the progression of the target function and monitor parameter coverage of the search space in both reduced 2D space and the full space. These visualisation can lead to insights for evaluating and comparing the performance of multiple optimisers in a specific problem. They can provide better understanding of existing methods or motivate the development of new approaches. For example, we have compared how three optimisers perform when maximising a non-smooth index function and illustrated how the pseudo-derivative search fails in this setting. The observations from our experiments have also been translated into improved optimisation methods for the guided tour, e.g. we introduced the option to interrupt the search if a better basis is found along the path.

>>>>

---
title: "Visual Diagnostics for Constrained Optimisation with Application to Guided Tours"
author:
  - name: H.Sherry Zhang
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics 
    url: Melbourne, Australia
    email:  huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    url: Melbourne, Australia
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    address: Institute of Statistics
    url: Vienna, Austria
    email:  ursula.laa@boku.ac.at  
  - name: Nicolas Langrené
    affiliation: CSIRO Data61 
    address: 34 Village Street, Docklands VIC 3008 Australia
    url: Melbourne, Australia
    email: nicolas.langrene@csiro.au
  - name: Patricia Menéndez
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    url: Melbourne, Australia
    email:  patricia.menendez@monash.edu 
abstract: >
  A guided tour helps to visualise high-dimensional data by showing low-dimensional projections along a projection pursuit optimisation path. Projection pursuit is a generalisation of principal component analysis, in the sense that different indexes are used to define the interestingness of the projected data. While much work has been done in developing new indexes in the literature, less has been done on understanding the optimisation. Index functions can be noisy, might have multiple local maxima as well as an optimal maximum, and are constrained to generate orthonormal projection frames, which complicates the optimization. In addition, projection pursuit is primarily used for exploratory data analysis, and finding the local maxima is also useful. The guided tour is especially useful for exploration, because it conducts geodesic interpolation connecting steps in the optimisation and shows how the projected data changes as a maxima is approached. This work provides new visual diagnostics for examining a choice of optimisation procedure, based on the provision of a new data object which collects information throughout the optimisation. It has helped to diagnose and fix several problems with projection pursuit guided tour. This work might be useful more broadly for diagnosing optimisers, and comparing their performance. The diagnostics are implemented in the R package [ferrn](https://github.com/huizezhang-sherry/ferrn).
keywords: graphics, multivariate, optimization
preamble: >
  \usepackage{mathtools}
  \usepackage[ruled, linesnumbered]{algorithm2e}
output: 
  rjtools::rjournal_web_article
bibliography: zhang-cook-laa-langrene-menendez.bib
---

@@

```{r toy-pca-animated-interactive,  out.wdith= "100px", fig.cap = "Animated version of Figure \\\\ref{fig:toy-pca}. With animation, the progression of the search paths from start to finish can better identified and CRS (green) finishes the optimisation quicker than PD (brown).", eval=knitr::is_html_output()}
```

@@

One interesting situation observed in the previous examples is that, for some simulations, as shown on the left panel of Figure \ref{fig:flip-sign}, the target basis is generated on the other half of the basis space and the interpolator seems to draw a straight line to interpolate. Bases with opposite signs do not affect the projection and index value, but we would prefer the target to have the same orientation as the current basis. The orientation of two bases can be checked via calculating the determinant and a negative value suggests the two bases having different orientation. For 1D bases, this can be corrected by flipping the sign in one of the basis while the correction for a general p-D bases is more complex and beyond the scope of this work. Here an orientation check is carried out once a new target basis is generated and the sign in the target basis will be flipped if a negative determinant is obtained. The interpolation after implementing the orientation check is shown on the right panel of Figure \ref{fig:flip-sign} where the unsatisfactory interpolation no longer exists.

@@

This paper has provided several visual diagnostics that can be used for understanding a complex optimisation procedure and are implemented in the ferrn package. The methods were illustrated using the optimisers available for projection pursuit guided tour. Here the constraint is the orthonormality condition of the projection bases. The approach used broadly applies to other constrained optimisers, by saving the path of the optimiser in a form required to input into the ferrn package. These high-dimensional data objects can then be viewed using the tour.

The progressive optimisation of a target function and it's coverage of the search space can be viewed in both reduced 2D space and the full space. These visualisation can lead to insights for evaluating and comparing the performance of multiple optimisers in a specific problem. They can provide better understanding of existing methods or motivate the development of new approaches. For example, we have compared how three optimisers perform when maximising a non-smooth index function and illustrated how the pseudo-derivative search fails in this setting. The observations from our experiments have also been translated into improved optimisation methods for the guided tour, e.g. we introduced the option to interrupt the search if a better basis is found along the path.

### 2021-09-16 paper/zhang-cook-laa-langrene-menendez.tex
changes made during the meeting

{\centering \includegraphics[width=0.7\linewidth]{/Users/sherryzhang/Documents/researh/paper-tour-vis/figs/tour-path}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/researh/paper-tour-vis/figs/toy-search}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/researh/paper-tour-vis/figs/toy-interp-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/researh/paper-tour-vis/figs/toy-pca}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/researh/paper-tour-vis/figs/toy-pca-animated-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/researh/paper-tour-vis/figs/toy-tour-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/researh/paper-tour-vis/figs/toy-torus-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/researh/paper-tour-vis/figs/interruption-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/researh/paper-tour-vis/figs/polish-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/researh/paper-tour-vis/figs/noisy-better-geo}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/researh/paper-tour-vis/figs/kol-result}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/researh/paper-tour-vis/figs/flip-sign}

@@

This paper has provided several visual diagnostics that can be used for
diagnosing a complex optimisation procedure. The methods were
illustrated using the optimisers available for projection pursuit guided
tour. Here the constraint is the orthonormality condition of the
projection bases. The approach used broadly applies to other constrained
optimisers.

The tools were implemented in the ferrn package and can track the
progression of the target function and monitor parameter coverage of the
search space in both reduced 2D space and the full space. These
visualisation can lead to insights for evaluating and comparing the
performance of multiple optimisers in a specific problem. They can
provide better understanding of existing methods or motivate the
development of new approaches. For example, we have compared how three
optimisers perform when maximising a non-smooth index function and
illustrated how the pseudo-derivative search fails in this setting. The
observations from our experiments have also been translated into
improved optimisation methods for the guided tour, e.g.~we introduced
the option to interrupt the search if a better basis is found along the
path.

>>>>

{\centering \includegraphics[width=0.7\linewidth]{/Users/sherryzhang/Documents/research/paper-tour-vis/figs/tour-path}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-tour-vis/figs/toy-search}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-tour-vis/figs/toy-interp-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-tour-vis/figs/toy-pca}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-tour-vis/figs/toy-pca-animated-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-tour-vis/figs/toy-tour-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-tour-vis/figs/toy-torus-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-tour-vis/figs/interruption-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-tour-vis/figs/polish-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-tour-vis/figs/noisy-better-geo}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-tour-vis/figs/kol-result}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-tour-vis/figs/flip-sign}

@@

This paper has provided several visual diagnostics that can be used for
understanding a complex optimisation procedure and are implemented in
the ferrn package. The methods were illustrated using the optimisers
available for projection pursuit guided tour. Here the constraint is the
orthonormality condition of the projection bases. The approach used
broadly applies to other constrained optimisers, by saving the path of
the optimiser in a form required to input into the ferrn package. These
high-dimensional data objects can then be viewed using the tour.

The progressive optimisation of a target function and it's coverage of
the search space can be viewed in both reduced 2D space and the full
space. These visualisation can lead to insights for evaluating and
comparing the performance of multiple optimisers in a specific problem.
They can provide better understanding of existing methods or motivate
the development of new approaches. For example, we have compared how
three optimisers perform when maximising a non-smooth index function and
illustrated how the pseudo-derivative search fails in this setting. The
observations from our experiments have also been translated into
improved optimisation methods for the guided tour, e.g.~we introduced
the option to interrupt the search if a better basis is found along the
path.
